---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Ropendal

Ropendal is an abstract remote filesystem operation interface for R backed by the Rust crate of [Apache OpenDAL](https://opendal.apache.org/). The R/Rust interface is built with the [`savvy`](https://github.com/yutannihilation/savvy) crate, which generates the R wrappers and native registration used by the package.

The central abstraction is an `OpendalFs` handle: a root-relative, byte-addressable filesystem over local files, HTTP endpoints, S3-compatible services, Google Drive, and other OpenDAL-backed providers as support is added. Operations move raw bytes and byte ranges, but metadata and namespace operations such as stat, exists, listing, delete, copy, rename, and mkdir are remote I/O too; backend failures resolve to classed error values; credentials are explicit rather than discovered through hidden ambient provider chains.

Aio handles are a core part of the interface. Reads, writes, metadata checks, listings, and namespace mutations can return nanonext-like handles that callers poll, collect, or cancel explicitly, while background Rust tasks never call R APIs. The package also exposes immutable `OpendalBytes` handles, chunked read/write iterators, paged listing/walking iterators for streaming-style transfer and namespace traversal, and installs `inst/include/ropendal.h`, a pure C API for downstream native packages that want async byte, metadata, and namespace operations.

See `design/api-design.md` for API design notes and `design/STATUS.md` for the implementation/test checklist.

## API at a glance

The tables below are generated from progress data frames in this README source.

```{r api-progress-tables, echo = FALSE, results = "asis"}
details_table <- function(title, data) {
  cat("<details>\n")
  cat("<summary>", title, "</summary>\n\n", sep = "")
  cat(knitr::kable(data, format = "pipe"), sep = "\n")
  cat("\n</details>\n\n")
}

operation_progress <- data.frame(
  operation = c("read", "write", "stat/exists", "list", "mkdir/delete", "copy", "rename", "append"),
  surface = c("sync, Aio, iterator", "sync, Aio, iterator", "sync, Aio", "sync, Aio, iterator", "sync, Aio", "sync, Aio", "sync, Aio", "sync, Aio, iterator"),
  status = c("implemented", "implemented", "implemented", "implemented", "implemented", "implemented", "backend-dependent", "backend-dependent"),
  notes = c(
    "range reads, batch concurrency, iterators, seek/tell, per-object read tuning",
    "create/replace, batch concurrency, iterators, tell, per-object write tuning",
    "metadata and existence values",
    "collectable or paged where provider supports listing",
    "root-relative path normalization",
    "direct provider copy",
    "no silent S3-style emulation of atomic rename",
    "returns unsupported where OpenDAL reports no append capability"
  )
)

aio_progress <- data.frame(
  abstraction = c("OpendalAio", "active bindings", "poll_aio()", "collect_aio()/call_aio()", "stop_aio()", "native C Aio"),
  status = c("implemented for bytes, metadata, entries, bools, and unit completions", "implemented", "implemented", "implemented", "implemented with delayed HTTP cancellation coverage", "implemented for byte, read-vector, metadata, entry/list, bool, and unit operations"),
  role = c(
    "nanonext-like handle for background Rust work",
    "$value/$data/$result plus $state/$resolved/$error",
    "non-blocking readiness check",
    "collect returns value; call waits/updates and returns the Aio invisibly",
    "explicit cancellation request",
    "downstream native packages can submit async filesystem operations"
  )
)

read_progress <- data.frame(
  function_or_path = c("fs_read()", "fs_read_aio()", "fs_read_bytes()/fs_read_bytes_aio()", "fs_read_iter()", "read_iter_next()", "read_iter_collect()", "fs_seek()/fs_tell()", "C read_aio()/read_into_aio()/readv_aio()/readv_into_aio()"),
  status = c("implemented", "implemented", "implemented", "implemented", "implemented", "implemented", "implemented", "implemented"),
  tuning = c(
    "batch_concurrency, read_concurrency, chunk_size, coalesce_gap",
    "same as fs_read()",
    "Rust-owned OpendalBytes handles; explicit as.raw() materialization",
    "one path returns a handle; many paths return a list of handles",
    "next chunk as raw bytes",
    "remaining chunks into one raw vector",
    "read iterator position within its read window",
    "borrowed result bytes or caller-owned output buffers"
  )
)

write_progress <- data.frame(
  function_or_path = c("fs_write()/fs_write_aio()", "fs_replace()/fs_replace_aio()", "fs_append()/fs_append_aio()", "fs_write_iter()", "write_iter_write()", "write_iter_close()", "fs_tell()", "C write/replace/append Aio"),
  status = c("implemented", "implemented", "backend-dependent", "implemented", "implemented", "implemented", "implemented", "implemented"),
  tuning = c(
    "batch_concurrency, write_concurrency, chunk_size; accepts raw or OpendalBytes",
    "same as fs_write(); accepts raw or OpendalBytes",
    "same where append is supported; accepts raw or OpendalBytes",
    "one path returns a sink; many paths return a list of sinks",
    "submit one raw chunk",
    "finalize multipart/streaming write",
    "bytes submitted to write sink; seek is intentionally unsupported",
    "caller-owned input buffer"
  )
)

provider_progress <- data.frame(
  provider = c("fs", "http", "s3-compatible", "gdrive", "gcs", "azblob", "other OpenDAL services"),
  status = c("implemented/tested", "implemented/tested read-only", "implemented/tested with public S3 and MinIO", "implemented/opt-in tested", "feature wired", "feature wired", "extension path"),
  credentials = c("none", "optional explicit headers", "explicit credentials_s3() or unsigned public config", "explicit credentials_gdrive()/credentials_gdrive3()", "planned explicit provider", "planned explicit provider", "needs config, credential, and tests")
)

serializer_progress <- data.frame(
  layer = c("raw bytes", "R serialize/unserialize", "text", "codecs/compression"),
  status = c("implemented", "implemented", "planned", "planned"),
  notes = c(
    "core storage contract uses R raw vectors and OpendalBytes handles",
    "mode = serial plus serial_config(), serialize_raw(), and deserialize_raw(); R hooks run on the R thread",
    "explicit encoding boundary",
    "separate from provider transfer chunking"
  )
)

details_table("Supported operations", operation_progress)
details_table("Aio interface", aio_progress)
details_table("Read operations", read_progress)
details_table("Write operations", write_progress)
details_table("Supported providers", provider_progress)
details_table("Serializers and codecs", serializer_progress)
```

## Installation

Source installs require a Rust toolchain. The default source build enables the currently wired OpenDAL service features for local filesystems, HTTP, S3-compatible storage, Google Cloud Storage, Azure Blob, and Google Drive. Since the core is backed by the OpenDAL Rust crate, Ropendal can grow to support additional OpenDAL services; adding a service still needs Ropendal-side configuration, credential, and test coverage.

```{r install-source-features, eval = FALSE}
# Keep only local filesystem, HTTP, S3-compatible, and Google Drive support.
install.packages(
  "Ropendal_0.0.0.9000.tar.gz",
  repos = NULL,
  type = "source",
  configure.args = "--without-default-rust-features --with-rust-features=fs,http,s3,gdrive"
)

# Add the current cloud-service feature group explicitly.
install.packages(
  "Ropendal_0.0.0.9000.tar.gz",
  repos = NULL,
  type = "source",
  configure.args = "--enable-cloud"
)
```

Equivalent environment-variable control is also supported for source builds.

```{bash install-source-env, eval = FALSE}
SAVVY_FEATURES="fs http s3 gdrive" R CMD INSTALL Ropendal_0.0.0.9000.tar.gz
```

## Local filesystem example

```{r local-fs-example}
library(Ropendal)

root <- file.path(tempdir(), "ropendal-readme-example")
unlink(root, recursive = TRUE)
dir.create(root, recursive = TRUE)

fs <- opendal("fs", root = root)

fs_write(fs, "data.bin", as.raw(c(1, 2, 3, 4)))
fs_read(fs, "data.bin")
fs_read(fs, "data.bin", offset = 1, size = 2)

fs_stat(fs, "data.bin")[c("path", "type", "size")]
bytes_handle <- fs_read_bytes(fs, "data.bin")
length(bytes_handle)
as.raw(bytes_handle)
vapply(fs_ls(fs), `[[`, character(1), "path")

it <- fs_ls_iter(fs, page_size = 1)
page <- ls_iter_next(it)
page$done
vapply(page$entries, `[[`, character(1), "path")
```

## Public S3-compatible endpoint example

Public S3-compatible endpoints can be opened without credentials by explicitly
skipping request signing and disabling ambient config loading. Private
S3-compatible services should use an explicit credential provider such as
`credentials_s3(access_key_id, secret_access_key)` rather than hidden ambient
credential lookup. This example uses the public 1000 Genomes AWS S3 bucket to
range-read the beginning of a large reference FASTA and inspect a small VCF
object without downloading an entire genome-scale file.

```{r public-s3-example, eval = FALSE}
# Public AWS S3 bucket — no credentials needed.
s3fs <- opendal(
  "s3",
  endpoint = "https://s3.amazonaws.com",
  bucket = "1000genomes",
  root = "/",
  region = "us-east-1",
  skip_signature = TRUE,
  disable_config_load = TRUE
)

fasta_path <- "technical/reference/GRCh38_reference_genome/GRCh38_full_analysis_set_plus_decoy_hla.fa"
fasta_head <- rawToChar(fs_read(s3fs, fasta_path, offset = 0, size = 80))
stopifnot(startsWith(fasta_head, ">chr1"))

fs_stat(s3fs, fasta_path)[c("path", "type", "size")]

vcf_path <- "phase3/integrated_sv_map/ALL.autosomes.pindel.20130502.complexindex.low_coverage.genotypes.vcf.gz"
fs_stat(s3fs, vcf_path)[c("path", "type", "size")]

vcf_head_gz <- fs_read(s3fs, vcf_path, offset = 0, size = 16384)
con <- gzcon(rawConnection(vcf_head_gz))
readLines(con, n = 3)
close(con)
```

HTTP endpoints are useful for read-only byte access too. Full reads and byte
ranges work when the server returns byte-range responses. For authenticated
HTTP(S) endpoints, pass explicit request headers such as `headers =
list(Authorization = "Bearer ...")`; Ropendal does not print these header
values.

```{r http-endpoint-example}
http_fs <- opendal(
  "http",
  endpoint = "https://uk1s3.embassy.ebi.ac.uk/idr/zarr/v0.4/idr0062A/6001240.zarr",
  root = "/"
)

zarray_head <- rawToChar(fs_read(http_fs, "0/.zarray", offset = 0, size = 40))
grepl('"chunks"', zarray_head, fixed = TRUE)

fs_stat(http_fs, "0/.zarray")[c("path", "type", "size")]

chunk_head <- fs_read(http_fs, "0/0/0/0/0", offset = 0, size = 16)
length(chunk_head)
```

## Google Drive example

Google Drive handles use explicit credentials. The README render does not supply
or discover credentials. To run this chunk locally, set `ROPENDAL_README_GDRIVE=true`
plus `ROPENDAL_GDRIVE_SECRET_JSON`, `ROPENDAL_GDRIVE_TOKENS_JSON`, and
`ROPENDAL_GDRIVE_ROOT` in the environment before rendering.

```{r gdrive-example, eval = identical(Sys.getenv("ROPENDAL_README_GDRIVE"), "true") && file.exists(Sys.getenv("ROPENDAL_GDRIVE_SECRET_JSON", "")) && file.exists(Sys.getenv("ROPENDAL_GDRIVE_TOKENS_JSON", ""))}
gdrive_fs <- opendal(
  "gdrive",
  root = Sys.getenv("ROPENDAL_GDRIVE_ROOT", "Ropendal"),
  auth = credentials_gdrive3(
    secret_json = Sys.getenv("ROPENDAL_GDRIVE_SECRET_JSON"),
    tokens_json = Sys.getenv("ROPENDAL_GDRIVE_TOKENS_JSON")
  )
)

gdrive_path <- Sys.getenv("ROPENDAL_GDRIVE_FILE", "map_catalog.txt")
fs_stat(gdrive_fs, gdrive_path)[c("path", "type", "size")]

catalog_head <- fs_read(gdrive_fs, gdrive_path, offset = 0, size = 80)
length(catalog_head)
```

Filesystem failures are returned as values.

```{r errors-as-values}
err <- fs_write(fs, "data.bin", as.raw(9))
is_error_value(err)
error_kind(err)
```

Remote operations can also return Aio handles.

```{r aio-read}
aio <- fs_read_aio(fs, "data.bin")
poll_aio(aio)
call_aio(aio)
aio$value

stat <- collect_aio(fs_stat_aio(fs, "data.bin"))
stat$path
stat$size
collect_aio(fs_exists_aio(fs, "data.bin"))
entry <- collect_aio(fs_ls_aio(fs))[[1]]
entry$path
entry$type
collect_aio(fs_replace_aio(fs, "data.bin", as.raw(c(4, 5, 6))))
```

## Related work

Apache OpenDAL supplies the Rust backend abstraction that Ropendal binds into R. The R [`objectstore`](https://github.com/Bisaloo/objectstore) package is related work for object-store access, and its [`store_filesystem.R`](https://github.com/Bisaloo/objectstore/blob/main/R/store_filesystem.R) implementation is a useful reference for filesystem-like surfaces in R. Ropendal takes a more opinionated byte-filesystem stance, with root-relative path normalization, explicit credentials, errors-as-values, async handles, and an exported native C API as part of the core contract.

The async API is inspired by [`nanonext`](https://github.com/shikokuchuo/nanonext)'s Aio model: operations return handles, callers can poll or wait explicitly, and notification primitives are designed so background workers never call R APIs.

## Development

Common development and opt-in service test targets are listed by `make help`.

```{bash make-help, comment = ""}
make help
```
