---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit this file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Ropendal

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/Ropendal/actions/workflows/r-package.yml/badge.svg)](https://github.com/sounkou-bioinfo/Ropendal/actions/workflows/r-package.yml)
[![R-universe](https://sounkou-bioinfo.r-universe.dev/badges/Ropendal)](https://sounkou-bioinfo.r-universe.dev/Ropendal)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Ropendal is an abstract filesystem interface for R backed by the Rust crate of
[Apache OpenDAL](https://opendal.apache.org/). The package is byte-first:
operations move raw bytes, then explicitly materialize into R objects through
modes and serializers.

The async API is inspired by `nanonext`-style Aio handles: issue async work,
then `call_aio()` to trigger execution and `collect_aio()` to retrieve results
(or an `opendalErrorValue`).

```{r badges, echo = FALSE}
# Keep this block intentionally minimal in README to avoid rendering noise.
```

## Installation

Ropendal is available from R-universe:

```{r install-default, eval = FALSE}
install.packages(
  "Ropendal",
  repos = c("https://sounkou-bioinfo.r-universe.dev", "https://cloud.r-project.org")
)
```

Source builds can still use explicit OpenDAL feature flags when you need custom
provider wiring.

```{r install-source-features, eval = FALSE}
# Keep only local filesystem, HTTP, S3-compatible, and Google Drive support.
install.packages(
  "Ropendal",
  repos = c("https://sounkou-bioinfo.r-universe.dev", "https://cloud.r-project.org"),
  type = "source",
  configure.args = "--without-default-rust-features --with-rust-features=fs,http,s3,gdrive"
)

# Add the current cloud-service feature group explicitly.
install.packages(
  "Ropendal",
  repos = c("https://sounkou-bioinfo.r-universe.dev", "https://cloud.r-project.org"),
  type = "source",
  configure.args = "--enable-cloud"
)
```

## Quick start

```{r quick-start}
library(Ropendal)

root <- file.path(tempdir(), "ropendal-readme")
unlink(root, recursive = TRUE)
dir.create(root, recursive = TRUE)

fs <- opendal("fs", root = root)

fs_write(fs, "note.txt", charToRaw("hello ropendal\n"))
raw <- fs_read(fs, "note.txt")
raw

stat <- fs_stat(fs, "note.txt")
stat[c("path", "type", "size")]

entries <- fs_ls(fs)
vapply(entries, `[[`, character(1), "path")
```

## Async reads and explicit wait

```{r quick-async}
aio <- fs_read_aio(fs, "note.txt")
call_aio(aio)
collect_aio(aio)
```

## Tour of the API across backends

The same `fs_*` calls apply to HTTP(S), S3, and Google Drive handles.

### HTTP/HTTPS read example

```{r api-tour-http, eval = FALSE}
http_fs <- opendal("http", endpoint = "https://raw.githubusercontent.com", root = "/")
http_head <- rawToChar(fs_read(http_fs, "sounkou-bioinfo/Ropendal/main/README.md", size = 64))
http_head
```

### Public S3-compatible read example

```{r api-tour-s3, eval = FALSE}
s3_fs <- opendal(
  "s3",
  endpoint = "https://uk1s3.embassy.ebi.ac.uk",
  bucket = "idr",
  root = "/zarr/v0.4/idr0062A/6001240.zarr",
  region = "us-east-1",
  skip_signature = TRUE,
  disable_config_load = TRUE
)

zarray <- rawToChar(fs_read(s3_fs, "0/.zarray", size = 64))
zarray
```

### Google Drive read example (credentials explicit)

```{r api-tour-gdrive, eval = FALSE}
drive_fs <- opendal(
  "gdrive",
  root = "Ropendal",
  auth = credentials_gdrive3(
    secret_json = "/path/to/secret.json",
    tokens_json = "/path/to/tokens.json",
    scope = "https://www.googleapis.com/auth/drive"
  )
)

drive_head <- rawToChar(fs_read(drive_fs, "map_catalog.txt", size = 64))
drive_head
```

## Development

Common targets:

```{bash make-help, comment = ""}
make --no-print-directory help
```
