---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit this file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Ropendal

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/Ropendal/actions/workflows/r-package.yml/badge.svg)](https://github.com/sounkou-bioinfo/Ropendal/actions/workflows/r-package.yml)
[![R-universe](https://sounkou-bioinfo.r-universe.dev/badges/Ropendal)](https://sounkou-bioinfo.r-universe.dev/Ropendal)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Ropendal: **Abstract Filesystem Access Interface for R** via the Rust crate
[`opendal`](https://crates.io/crates/opendal), using
[`savvy`](https://github.com/yutannihilation/savvy) for R/Rust FFI. We keep the
bottom layer byte-first: filesystem operations move raw bytes, and serializers
or codecs explicitly materialize those bytes into R objects.

Async work uses Aio handles inspired by
[`nanonext`](https://nanonext.r-lib.org/): we issue work with `fs_*_aio()`, wait
with `call_aio()`, and retrieve the payload (or an `opendalErrorValue`) with
`collect_aio()`.

We expose the same principles to native packages through a pure C API in
`inst/include/ropendal.h`: submit async work, wait on an Aio, then read borrowed
results or fill caller-owned buffers without routing data through R raw vectors.

Backends are configured explicitly. We pass HTTP headers to `opendal("http")`
with `headers`, and we pass credential provider objects such as
`credentials_s3()`, `credentials_gcs()`, `credentials_azblob()`, and
`credentials_gdrive()` with `auth`.

```{r badges, echo = FALSE}
# Keep this block intentionally minimal in README to avoid rendering noise.
```

## Installation

Ropendal is available from R-universe:

```{r install-default, eval = FALSE}
install.packages(
  "Ropendal",
  repos = c("https://sounkou-bioinfo.r-universe.dev", "https://cloud.r-project.org")
)
```

Source builds can still use explicit OpenDAL feature flags when you need custom
provider wiring.

```{r install-source-features, eval = FALSE}
# Keep only local filesystem, HTTP, S3-compatible, and Google Drive support.
install.packages(
  "Ropendal",
  repos = c("https://sounkou-bioinfo.r-universe.dev", "https://cloud.r-project.org"),
  type = "source",
  configure.args = "--without-default-rust-features --with-rust-features=fs,http,s3,gdrive"
)

# Add the current cloud-service feature group explicitly.
install.packages(
  "Ropendal",
  repos = c("https://sounkou-bioinfo.r-universe.dev", "https://cloud.r-project.org"),
  type = "source",
  configure.args = "--enable-cloud"
)
```

## Quick start

```{r quick-start}
library(Ropendal)

root <- file.path(tempdir(), "ropendal-readme")
unlink(root, recursive = TRUE)
dir.create(root, recursive = TRUE)

fs <- opendal("fs", root = root)

fs_write(fs, "note.txt", charToRaw("hello ropendal\n"))
raw <- fs_read(fs, "note.txt")
rawToChar(raw)

stat <- fs_stat(fs, "note.txt")
stat[c("path", "type", "size")]

entries <- fs_ls(fs)
vapply(entries, `[[`, character(1), "path")
```

## Async reads and explicit wait

We get an Aio handle, wait for it, and collect the result when needed. `fs_read()`
is vectorized too: we can pass several paths with `batch_concurrency` when we
want one call to launch multiple reads at the same time.

```{r quick-async}
aio <- fs_read_aio(fs, "note.txt")
call_aio(aio)
collect_aio(aio)
```

## Tour of the API across backends

The same `fs_*` calls apply to HTTP(S), S3, and Google Drive handles.

### HTTP/HTTPS read example

We define the headers once, require them on the local fixture, and pass the same
headers to the HTTP filesystem handle.

```{r api-tour-http}
root <- tempfile("ropendal-http-readme-")
dir.create(root, recursive = TRUE)
writeLines("hello http example", file.path(root, "hello.txt"))

headers <- list(
  Authorization = "Bearer ropendal-readme",
  `X-Ropendal-Example` = "headers"
)

fixture <- OpendalHttpFixture$start(root, required_headers = headers)
http_fs <- opendal("http", endpoint = fixture$endpoint(), root = "/", headers = headers)

rawToChar(fs_read(http_fs, "hello.txt"))
fs_stat(http_fs, "hello.txt")[c("path", "type", "size")]

fixture$stop()
```

```{r readme-minio-helper, include = FALSE}
readme_run_s3 <- requireNamespace("processx", quietly = TRUE) &&
  requireNamespace("bench", quietly = TRUE) &&
  requireNamespace("paws.storage", quietly = TRUE) &&
  nchar(Sys.which("minio")) > 0L &&
  nchar(Sys.which("mc")) > 0L

readme_start_minio <- function() {
  minio_bin <- Sys.which("minio")
  mc_bin <- Sys.which("mc")
  stopifnot(nchar(minio_bin) > 0L, nchar(mc_bin) > 0L)

  ports <- sample(11000:64000, 2L)
  endpoint <- sprintf("http://127.0.0.1:%s", ports[[1L]])
  data_dir <- tempfile("ropendal-minio-readme-data-")
  dir.create(data_dir, recursive = TRUE)
  log_file <- tempfile("ropendal-minio-readme-log-")

  bucket <- "ropendal-readme"
  access_key <- "minioadmin"
  secret_key <- paste0(sample(c(letters, LETTERS, 0:9), 24L, replace = TRUE), collapse = "")
  alias <- paste0("ropendal-readme-", Sys.getpid())

  minio_env <- Sys.getenv()
  minio_env["MINIO_ROOT_USER"] <- access_key
  minio_env["MINIO_ROOT_PASSWORD"] <- secret_key

  minio_proc <- processx::process$new(
    command = minio_bin,
    args = c(
      "server",
      data_dir,
      "--address",
      sprintf("127.0.0.1:%s", ports[[1L]]),
      "--console-address",
      sprintf("127.0.0.1:%s", ports[[2L]])
    ),
    env = minio_env,
    stdout = log_file,
    stderr = log_file
  )

  stop_minio <- function() {
    try(processx::run(mc_bin, c("alias", "remove", alias), error_on_status = FALSE), silent = TRUE)
    if (minio_proc$is_alive()) minio_proc$kill()
    minio_proc$wait()
    unlink(data_dir, recursive = TRUE)
    unlink(log_file)
  }

  ready <- FALSE
  for (attempt in seq_len(80)) {
    alias_result <- try(
      processx::run(
        command = mc_bin,
        args = c("alias", "set", alias, endpoint, access_key, secret_key),
        error_on_status = FALSE
      ),
      silent = TRUE
    )
    ready <- !inherits(alias_result, "try-error") && identical(alias_result$status, 0L)
    if (ready) break
    if (!minio_proc$is_alive()) {
      details <- if (file.exists(log_file)) paste(readLines(log_file, warn = FALSE), collapse = "\n") else ""
      stop("minio exited before it was ready\n", details, call. = FALSE)
    }
    Sys.sleep(0.25)
  }
  if (!ready) stop("minio did not become ready", call. = FALSE)

  processx::run(mc_bin, c("mb", "--ignore-existing", paste0(alias, "/", bucket)), error_on_status = TRUE)

  auth <- credentials_s3(
    access_key_id = access_key,
    secret_access_key = secret_key,
    region = "us-east-1",
    source = "readme-minio"
  )

  fs <- opendal(
    "s3",
    endpoint = endpoint,
    bucket = bucket,
    root = "/",
    disable_config_load = TRUE,
    disable_ec2_metadata = TRUE,
    auth = auth
  )

  list(
    fs = fs,
    stop = stop_minio,
    endpoint = endpoint,
    bucket = bucket,
    access_key = access_key,
    secret_key = secret_key,
    region = "us-east-1"
  )
}

readme_set_aws_env <- function(minio) {
  old <- Sys.getenv(c("AWS_ACCESS_KEY_ID", "AWS_SECRET_ACCESS_KEY", "AWS_REGION"), unset = NA)
  Sys.setenv(
    AWS_ACCESS_KEY_ID = minio$access_key,
    AWS_SECRET_ACCESS_KEY = minio$secret_key,
    AWS_REGION = minio$region
  )
  function() {
    for (nm in names(old)) {
      if (is.na(old[[nm]])) Sys.unsetenv(nm) else do.call(Sys.setenv, as.list(stats::setNames(old[[nm]], nm)))
    }
  }
}
```

### S3-compatible store (local MinIO)

We set up a temporary MinIO instance behind the scenes, then check that the
S3-compatible store supports the same byte API.

```{r api-tour-s3-start, include = FALSE, eval = readme_run_s3}
minio <- readme_start_minio()
s3_fs <- minio$fs
```

We write, read, stat, and list objects with the same `fs_*` functions.

```{r api-tour-s3-ops, eval = readme_run_s3}
fs_write(s3_fs, "notes/a.txt", charToRaw("hello s3-compatible store\n"))
fs_write(s3_fs, "notes/b.txt", charToRaw("another object\n"))

rawToChar(fs_read(s3_fs, "notes/a.txt"))
fs_stat(s3_fs, "notes/a.txt")[c("path", "type", "size")]
vapply(fs_ls(s3_fs, "notes/"), `[[`, character(1), "path")
```

The async path returns an Aio handle; we wait and collect when we need the
payload.

```{r api-tour-s3-async, eval = readme_run_s3}
aio <- fs_read_aio(s3_fs, "notes/b.txt")
call_aio(aio)
rawToChar(collect_aio(aio))
```

For a small local comparison, we use a larger object to compare Ropendal's
default path with its chunked/concurrent path. `paws.storage` is included as a
single-GET baseline.

```{r api-tour-s3-bench-setup, eval = readme_run_s3}
restore_aws_env <- readme_set_aws_env(minio)
paws_s3 <- paws.storage::s3(
  endpoint = minio$endpoint,
  region = minio$region,
  config = list(s3_force_path_style = TRUE)
)

payload <- as.raw(sample.int(256L, 8L * 1024L * 1024L, replace = TRUE) - 1L)
bench_key <- "bench/payload.bin"
read_chunk <- 1024 * 1024
write_chunk <- 5 * 1024 * 1024
```

We first compare upload/replace paths: the default Ropendal call, Ropendal with
explicit write chunking/concurrency, and `paws.storage::put_object()`.

```{r api-tour-s3-bench-write, eval = readme_run_s3}
bench::mark(
  ropendal_replace = fs_replace(s3_fs, bench_key, payload),
  ropendal_replace_concurrent = fs_replace(
    s3_fs,
    bench_key,
    payload,
    write_concurrency = 4,
    chunk_size = write_chunk
  ),
  paws_put = {
    paws_s3$put_object(Bucket = minio$bucket, Key = bench_key, Body = payload)
    TRUE
  },
  iterations = 3,
  check = FALSE
)[, c("expression", "min", "median", "itr/sec", "mem_alloc", "n_gc")]
```

Then we compare download paths. The Ropendal rows separate default reads,
chunked/concurrent reads, Aio reads, and Aio plus chunked/concurrent reads;
`paws_get` remains the single-GET baseline.

```{r api-tour-s3-bench-read, eval = readme_run_s3}
bench::mark(
  ropendal_read = fs_read(s3_fs, bench_key),
  ropendal_read_concurrent = fs_read(
    s3_fs,
    bench_key,
    read_concurrency = 4,
    chunk_size = read_chunk
  ),
  ropendal_read_aio = collect_aio(fs_read_aio(s3_fs, bench_key)),
  ropendal_read_aio_concurrent = collect_aio(fs_read_aio(
    s3_fs,
    bench_key,
    read_concurrency = 4,
    chunk_size = read_chunk
  )),
  paws_get = paws_s3$get_object(Bucket = minio$bucket, Key = bench_key)$Body,
  iterations = 3,
  check = FALSE
)[, c("expression", "min", "median", "itr/sec", "mem_alloc", "n_gc")]
```

```{r api-tour-s3-cleanup, include = FALSE, eval = readme_run_s3}
restore_aws_env()
minio$stop()
```

### Google Drive read example (credentials explicit)

For Google Drive, we pass a credential provider object through `auth` and keep
secret material outside the filesystem handle printout.

```{r api-tour-gdrive, eval = identical(Sys.getenv("ROPENDAL_README_GDRIVE", "false"), "true")}
secret_json <- Sys.getenv("ROPENDAL_GDRIVE_SECRET_JSON")
tokens_json <- Sys.getenv("ROPENDAL_GDRIVE_TOKENS_JSON", unset = file.path(dirname(secret_json), "tokens.json"))
gdrive_root <- Sys.getenv("ROPENDAL_GDRIVE_ROOT", unset = "Ropendal")
gdrive_file <- Sys.getenv("ROPENDAL_GDRIVE_FILE", unset = "map_catalog.txt")

drive_fs <- opendal(
  "gdrive",
  root = gdrive_root,
  auth = credentials_gdrive3(
    secret_json = secret_json,
    tokens_json = tokens_json,
    scope = "https://www.googleapis.com/auth/drive"
  )
)
drive_head <- rawToChar(fs_read(drive_fs, gdrive_file, size = 64))
drive_head
```

## Native C API smoke test

The native API is for other R packages that want OpenDAL-backed byte I/O without
calling R while async work is running. A downstream package can declare
`LinkingTo: Ropendal`, include `<ropendal.h>`, submit async work, wait on Aio
handles, and read borrowed results or fill caller-owned buffers.

We exercise that installed C API in-process with
[`Rtinycc`](https://github.com/sounkou-bioinfo/Rtinycc). The C code submits
async work, waits on Aio handles, checks metadata, lists entries, and reads into
a caller-owned buffer.

```{r native-c-api-rtinycc, eval = requireNamespace("Rtinycc", quietly = TRUE)}
root <- tempfile("ropendal-c-api-readme-")
dir.create(root, recursive = TRUE)

ropendal_lib <- list.files(
  system.file("libs", package = "Ropendal"),
  pattern = paste0("Ropendal", .Platform$dynlib.ext, "$"),
  recursive = TRUE,
  full.names = TRUE
)

c_api_code <- '
#include <stdint.h>
#include <string.h>
#include "ropendal.h"

static int cleanup(ropendal_fs_t *fs, ropendal_aio_t *aio, ropendal_error_t *err) {
  if (aio) ropendal_aio_release(aio);
  if (fs) ropendal_fs_release(fs);
  if (err) ropendal_error_release(err);
  return -1;
}

int ropendal_c_api_smoke(const char *root) {
  ropendal_error_t *err = 0;
  ropendal_fs_t *fs = 0;
  ropendal_aio_t *aio = 0;
  ropendal_kv_t cfg = { sizeof(ropendal_kv_t), "root", root };

  ropendal_status_t st = ropendal_fs_open("fs", &cfg, 1, &fs, &err);
  if (st != ROPENDAL_OK) return cleanup(fs, aio, err);

  const uint8_t payload[] = "hello native api\\n";
  const size_t payload_len = sizeof(payload) - 1;
  ropendal_write_options_t w = {0};
  w.struct_size = sizeof(w);
  w.path = "native.txt";

  st = ropendal_write_aio(fs, &w, payload, payload_len, &aio, &err);
  if (st != ROPENDAL_OK) return cleanup(fs, aio, err);
  st = ropendal_aio_wait(aio, -1, &err);
  if (st != ROPENDAL_OK) return cleanup(fs, aio, err);
  ropendal_aio_release(aio);
  aio = 0;

  st = ropendal_stat_aio(fs, "native.txt", 0, 0, &aio, &err);
  if (st != ROPENDAL_OK) return cleanup(fs, aio, err);
  st = ropendal_aio_wait(aio, -1, &err);
  if (st != ROPENDAL_OK) return cleanup(fs, aio, err);
  const ropendal_entry_t *entry = 0;
  st = ropendal_aio_result_entry(aio, &entry, &err);
  if (st != ROPENDAL_OK || !entry || !entry->has_content_length || entry->content_length != payload_len) {
    return cleanup(fs, aio, err);
  }
  ropendal_aio_release(aio);
  aio = 0;

  ropendal_ls_options_t ls = {0};
  ls.struct_size = sizeof(ls);
  ls.path = "";
  st = ropendal_ls_aio(fs, &ls, &aio, &err);
  if (st != ROPENDAL_OK) return cleanup(fs, aio, err);
  st = ropendal_aio_wait(aio, -1, &err);
  if (st != ROPENDAL_OK) return cleanup(fs, aio, err);
  const ropendal_entry_t *entries = 0;
  size_t nentries = 0;
  st = ropendal_aio_result_entries(aio, &entries, &nentries, &err);
  if (st != ROPENDAL_OK || nentries == 0) return cleanup(fs, aio, err);
  ropendal_aio_release(aio);
  aio = 0;

  uint8_t dst[64] = {0};
  ropendal_read_options_t r = {0};
  r.struct_size = sizeof(r);
  r.path = "native.txt";
  st = ropendal_read_into_aio(fs, &r, dst, sizeof(dst), &aio, &err);
  if (st != ROPENDAL_OK) return cleanup(fs, aio, err);
  st = ropendal_aio_wait(aio, -1, &err);
  if (st != ROPENDAL_OK) return cleanup(fs, aio, err);
  size_t nread = 0;
  st = ropendal_aio_result_nread(aio, &nread, &err);
  if (st != ROPENDAL_OK || nread != payload_len || memcmp(dst, payload, payload_len) != 0) {
    return cleanup(fs, aio, err);
  }

  ropendal_aio_release(aio);
  ropendal_fs_release(fs);
  return (int)nread;
}
'

ffi <- Rtinycc::tcc_ffi() |>
  Rtinycc::tcc_include(system.file("include", package = "Ropendal")) |>
  Rtinycc::tcc_library(ropendal_lib[[1]]) |>
  Rtinycc::tcc_source(c_api_code) |>
  Rtinycc::tcc_bind(
    ropendal_c_api_smoke = list(args = list("cstring"), returns = "i32")
  ) |>
  Rtinycc::tcc_compile()

ffi$ropendal_c_api_smoke(root)
```

## Development

Common targets:

```{bash make-help, comment = ""}
make --no-print-directory help
```
