---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit this file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Ropendal

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/Ropendal/actions/workflows/r-package.yml/badge.svg)](https://github.com/sounkou-bioinfo/Ropendal/actions/workflows/r-package.yml)
[![R-universe](https://sounkou-bioinfo.r-universe.dev/badges/Ropendal)](https://sounkou-bioinfo.r-universe.dev/Ropendal)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Ropendal: **Abstract Filesystem Access Interface for R** via the Rust crate
[`opendal`](https://crates.io/crates/opendal), using
[`savvy`](https://github.com/yutannihilation/savvy) for R/Rust FFI. We keep the
bottom layer byte-first: filesystem operations move raw bytes, and serializers
or codecs explicitly materialize those bytes into R objects.

Async work uses Aio handles inspired by
[`nanonext`](https://nanonext.r-lib.org/): we issue work with `fs_*_aio()`, wait
with `call_aio()`, and retrieve the payload (or an `opendalErrorValue`) with
`collect_aio()`.

We expose the same principles to native packages through a pure C API in
`inst/include/ropendal.h`: submit async work, wait on an Aio, then read borrowed
results or fill caller-owned buffers without routing data through R raw vectors.

Serializers and codecs are first-class. Serializer/deserializer pairs control
R-object materialization; native codecs such as `gzip` and `zlib` transform raw
bytes. Both layers stay explicit through `mode`, `serial_config()`,
`codec_config()`, `serialize_raw()`, and `deserialize_raw()`.

Backends are configured explicitly. We pass HTTP headers to `opendal("http")`
with `headers`, and we pass credential provider objects such as
`credentials_s3()`, `credentials_gcs()`, `credentials_azblob()`, and
`credentials_gdrive()` with `auth`.

```{r badges, echo = FALSE}
# Keep this block intentionally minimal in README to avoid rendering noise.
```

## Installation

Ropendal is available from R-universe:

```{r install-default, eval = FALSE}
install.packages(
  "Ropendal",
  repos = c("https://sounkou-bioinfo.r-universe.dev", "https://cloud.r-project.org")
)
```

R-universe also publishes Linux binaries for Ubuntu `noble` on `x86_64` and
`arm64` for R-release and R-devel. To prefer those binaries on Linux, point
`repos` at the known binary repositories for this universe and CRAN dependencies:

```{r install-linux-binaries, eval = FALSE}
options(repos = c(
  Ropendal = sprintf(
    "https://sounkou-bioinfo.r-universe.dev/bin/linux/noble-%s/%s/",
    R.version$arch,
    substr(getRversion(), 1, 3)
  ),
  CRAN = sprintf(
    "https://cran.r-universe.dev/bin/linux/noble-%s/%s/",
    R.version$arch,
    substr(getRversion(), 1, 3)
  )
))
install.packages("Ropendal")
```

Source builds can still use explicit OpenDAL feature flags when you need custom
provider wiring.

```{r install-source-features, eval = FALSE}
# Keep only local filesystem, HTTP, S3-compatible, and Google Drive support.
install.packages(
  "Ropendal",
  repos = c("https://sounkou-bioinfo.r-universe.dev", "https://cloud.r-project.org"),
  type = "source",
  configure.args = "--without-default-rust-features --with-rust-features=fs,http,s3,gdrive"
)

# Add the current cloud-service feature group explicitly.
install.packages(
  "Ropendal",
  repos = c("https://sounkou-bioinfo.r-universe.dev", "https://cloud.r-project.org"),
  type = "source",
  configure.args = "--enable-cloud"
)
```

## A Quick Start

A single filesystem handle gives us byte primitives, vectorized batches, Aio
handles, explicit serializers/codecs, and lower-level iterators.

### Create a handle and move bytes

```{r quick-start-handle}
library(Ropendal)

root <- tempfile("ropendal-readme-")
dir.create(root, recursive = TRUE)

fs <- opendal("fs", root = root)

fs_write(fs, "note.txt", charToRaw("hello ropendal\n"))
rawToChar(fs_read(fs, "note.txt"))
fs_stat(fs, "note.txt")[c("path", "type", "size")]
vapply(fs_ls(fs), `[[`, character(1), "path")
```

### Submit a batch in one call

```{r quick-start-batch}
paths <- c("batch/one.txt", "batch/two.txt")
fs_write(
  fs,
  paths,
  list(charToRaw("one\n"), charToRaw("two\n")),
  batch_concurrency = 2
)

many <- fs_read(fs, paths, offset = c(0, 0), batch_concurrency = 2)
vapply(many, rawToChar, character(1))
```

### Start async work and collect later

```{r quick-start-async}
aio <- fs_read_aio(fs, "note.txt")
call_aio(aio)
rawToChar(collect_aio(aio))
```

### Materialize tables and bytes explicitly

A serializer/deserializer can use
[`nanoarrow`](https://arrow.apache.org/nanoarrow/latest/r/) to turn a data frame
into Arrow IPC bytes and back. Native codecs such as `gzip` remain explicit byte
transforms.

```{r quick-start-materializers, eval = requireNamespace("nanoarrow", quietly = TRUE)}
arrow_config <- serial_config(
  "data.frame",
  sfunc = function(x) {
    con <- rawConnection(raw(), "wb")
    on.exit(close(con), add = TRUE)
    nanoarrow::write_nanoarrow(x, con)
    rawConnectionValue(con)
  },
  ufunc = function(raw) as.data.frame(nanoarrow::read_nanoarrow(raw))
)

arrow_tbl <- data.frame(
  id = 1:3,
  sample = c("HG001", "HG002", "HG003"),
  depth = c(32.5, 28.0, 41.25)
)

fs_replace(fs, "tables/depth.arrows", arrow_tbl, mode = "serial", serial_config = arrow_config)
fs_read(fs, "tables/depth.arrows", mode = "serial", serial_config = arrow_config)

fs_replace(fs, "objects/message.gz", charToRaw("compressed bytes\n"), mode = "codec", codec = "gzip")
rawToChar(fs_read(fs, "objects/message.gz", mode = "codec", codec = "gzip"))
```

### Drop down to lower-level byte iterators

```{r quick-start-iterators}
writer <- fs_write_iter(fs, "stream.txt")
write_iter_write(writer, charToRaw("hello "))
write_iter_write(writer, charToRaw("iterator\n"))
write_iter_close(writer)

reader <- fs_read_iter(fs, "stream.txt", chunk_size = 5)
first <- read_iter_next(reader)
rawToChar(first$data)
fs_tell(reader)
fs_seek(reader, 6)
rawToChar(read_iter_collect(reader))
```

## Tour of the API across backends

The same `fs_*` calls apply to HTTP(S), S3, and Google Drive handles.

### HTTP/HTTPS read example

We define the headers once, require them on the local fixture, and pass the same
headers to the HTTP filesystem handle.

```{r api-tour-http}
root <- tempfile("ropendal-http-readme-")
dir.create(root, recursive = TRUE)
writeLines("hello http example", file.path(root, "hello.txt"))

headers <- list(
  Authorization = "Bearer ropendal-readme",
  `X-Ropendal-Example` = "headers"
)

fixture <- OpendalHttpFixture$start(root, required_headers = headers)
http_fs <- opendal("http", endpoint = fixture$endpoint(), root = "/", headers = headers)

rawToChar(fs_read(http_fs, "hello.txt"))
fs_stat(http_fs, "hello.txt")[c("path", "type", "size")]

fixture$stop()
```

```{r readme-minio-helper, include = FALSE}
readme_run_s3 <- requireNamespace("processx", quietly = TRUE) &&
  requireNamespace("bench", quietly = TRUE) &&
  requireNamespace("paws.storage", quietly = TRUE) &&
  nchar(Sys.which("minio")) > 0L &&
  nchar(Sys.which("mc")) > 0L

readme_start_minio <- function() {
  minio_bin <- Sys.which("minio")
  mc_bin <- Sys.which("mc")
  stopifnot(nchar(minio_bin) > 0L, nchar(mc_bin) > 0L)

  ports <- sample(11000:64000, 2L)
  endpoint <- sprintf("http://127.0.0.1:%s", ports[[1L]])
  data_dir <- tempfile("ropendal-minio-readme-data-")
  dir.create(data_dir, recursive = TRUE)
  log_file <- tempfile("ropendal-minio-readme-log-")

  bucket <- "ropendal-readme"
  access_key <- "minioadmin"
  secret_key <- paste0(sample(c(letters, LETTERS, 0:9), 24L, replace = TRUE), collapse = "")
  alias <- paste0("ropendal-readme-", Sys.getpid())

  minio_env <- Sys.getenv()
  minio_env["MINIO_ROOT_USER"] <- access_key
  minio_env["MINIO_ROOT_PASSWORD"] <- secret_key

  minio_proc <- processx::process$new(
    command = minio_bin,
    args = c(
      "server",
      data_dir,
      "--address",
      sprintf("127.0.0.1:%s", ports[[1L]]),
      "--console-address",
      sprintf("127.0.0.1:%s", ports[[2L]])
    ),
    env = minio_env,
    stdout = log_file,
    stderr = log_file
  )

  stop_minio <- function() {
    try(processx::run(mc_bin, c("alias", "remove", alias), error_on_status = FALSE), silent = TRUE)
    if (minio_proc$is_alive()) minio_proc$kill()
    minio_proc$wait()
    unlink(data_dir, recursive = TRUE)
    unlink(log_file)
  }

  ready <- FALSE
  for (attempt in seq_len(80)) {
    alias_result <- try(
      processx::run(
        command = mc_bin,
        args = c("alias", "set", alias, endpoint, access_key, secret_key),
        error_on_status = FALSE
      ),
      silent = TRUE
    )
    ready <- !inherits(alias_result, "try-error") && identical(alias_result$status, 0L)
    if (ready) break
    if (!minio_proc$is_alive()) {
      details <- if (file.exists(log_file)) paste(readLines(log_file, warn = FALSE), collapse = "\n") else ""
      stop("minio exited before it was ready\n", details, call. = FALSE)
    }
    Sys.sleep(0.25)
  }
  if (!ready) stop("minio did not become ready", call. = FALSE)

  processx::run(mc_bin, c("mb", "--ignore-existing", paste0(alias, "/", bucket)), error_on_status = TRUE)

  auth <- credentials_s3(
    access_key_id = access_key,
    secret_access_key = secret_key,
    region = "us-east-1",
    source = "readme-minio"
  )

  fs <- opendal(
    "s3",
    endpoint = endpoint,
    bucket = bucket,
    root = "/",
    disable_config_load = TRUE,
    disable_ec2_metadata = TRUE,
    auth = auth
  )

  list(
    fs = fs,
    stop = stop_minio,
    endpoint = endpoint,
    bucket = bucket,
    access_key = access_key,
    secret_key = secret_key,
    region = "us-east-1"
  )
}

readme_set_aws_env <- function(minio) {
  old <- Sys.getenv(c("AWS_ACCESS_KEY_ID", "AWS_SECRET_ACCESS_KEY", "AWS_REGION"), unset = NA)
  Sys.setenv(
    AWS_ACCESS_KEY_ID = minio$access_key,
    AWS_SECRET_ACCESS_KEY = minio$secret_key,
    AWS_REGION = minio$region
  )
  function() {
    for (nm in names(old)) {
      if (is.na(old[[nm]])) Sys.unsetenv(nm) else do.call(Sys.setenv, as.list(stats::setNames(old[[nm]], nm)))
    }
  }
}
```

### S3-compatible store (local MinIO)

We set up a temporary MinIO instance behind the scenes, then check that the
S3-compatible store supports the same byte API.

```{r api-tour-s3-start, include = FALSE, eval = readme_run_s3}
minio <- readme_start_minio()
s3_fs <- minio$fs
```

We write, read, stat, and list objects with the same `fs_*` functions.

```{r api-tour-s3-ops, eval = readme_run_s3}
fs_write(s3_fs, "notes/a.txt", charToRaw("hello s3-compatible store\n"))
fs_write(s3_fs, "notes/b.txt", charToRaw("another object\n"))

rawToChar(fs_read(s3_fs, "notes/a.txt"))
fs_stat(s3_fs, "notes/a.txt")[c("path", "type", "size")]
vapply(fs_ls(s3_fs, "notes/"), `[[`, character(1), "path")
```

The async path returns an Aio handle; we wait and collect when we need the
payload.

```{r api-tour-s3-async, eval = readme_run_s3}
aio <- fs_read_aio(s3_fs, "notes/b.txt")
call_aio(aio)
rawToChar(collect_aio(aio))
```

For a small local comparison, we use a larger object to compare Ropendal's
default path with its chunked/concurrent path. Local MinIO removes most network
latency, so concurrency helps most when the object is large enough or when the
store is remote; for many independent objects, use `batch_concurrency`.
`paws.storage` is included as a single-GET baseline.

```{r api-tour-s3-bench-setup, eval = readme_run_s3}
restore_aws_env <- readme_set_aws_env(minio)
paws_s3 <- paws.storage::s3(
  endpoint = minio$endpoint,
  region = minio$region,
  config = list(s3_force_path_style = TRUE)
)

payload_size <- 64 * 1024 * 1024
payload <- rep(as.raw(0:255), length.out = payload_size)
bench_key <- "bench/payload.bin"
read_chunk <- 8 * 1024 * 1024
write_chunk <- 8 * 1024 * 1024
```

We first compare upload/replace paths: the default Ropendal call, Ropendal with
explicit write chunking/concurrency, and `paws.storage::put_object()`.

```{r api-tour-s3-bench-write, eval = readme_run_s3}
bench::mark(
  ropendal_replace = fs_replace(s3_fs, bench_key, payload),
  ropendal_replace_concurrent = fs_replace(
    s3_fs,
    bench_key,
    payload,
    write_concurrency = 4,
    chunk_size = write_chunk
  ),
  paws_put = {
    paws_s3$put_object(Bucket = minio$bucket, Key = bench_key, Body = payload)
    TRUE
  },
  iterations = 3,
  check = FALSE
)[, c("expression", "min", "median", "itr/sec", "mem_alloc", "n_gc")]
```

Then we compare download paths. The Ropendal rows separate default reads,
chunked/concurrent reads, Aio reads, and Aio plus chunked/concurrent reads;
`paws_get` remains the single-GET baseline.

```{r api-tour-s3-bench-read, eval = readme_run_s3}
bench::mark(
  ropendal_read = fs_read(s3_fs, bench_key),
  ropendal_read_concurrent = fs_read(
    s3_fs,
    bench_key,
    read_concurrency = 4,
    chunk_size = read_chunk
  ),
  ropendal_read_aio = collect_aio(fs_read_aio(s3_fs, bench_key)),
  ropendal_read_aio_concurrent = collect_aio(fs_read_aio(
    s3_fs,
    bench_key,
    read_concurrency = 4,
    chunk_size = read_chunk
  )),
  paws_get = paws_s3$get_object(Bucket = minio$bucket, Key = bench_key)$Body,
  iterations = 3,
  check = FALSE
)[, c("expression", "min", "median", "itr/sec", "mem_alloc", "n_gc")]
```

```{r api-tour-s3-cleanup, include = FALSE, eval = readme_run_s3}
restore_aws_env()
minio$stop()
```

### Google Drive read example (credentials explicit)

For Google Drive, we pass a credential provider object through `auth` and keep
secret material outside the filesystem handle printout. The chunk evaluates when
local credential files are available.

```{r readme-gdrive-helper, include = FALSE}
readme_gdrive_secret_json <- Sys.getenv("ROPENDAL_GDRIVE_SECRET_JSON")
readme_gdrive_tokens_json <- Sys.getenv(
  "ROPENDAL_GDRIVE_TOKENS_JSON",
  unset = if (nzchar(readme_gdrive_secret_json)) file.path(dirname(readme_gdrive_secret_json), "tokens.json") else ""
)
readme_run_gdrive <- nzchar(readme_gdrive_secret_json) &&
  nzchar(readme_gdrive_tokens_json) &&
  file.exists(readme_gdrive_secret_json) &&
  file.exists(readme_gdrive_tokens_json)
```

```{r api-tour-gdrive, eval = readme_run_gdrive}
secret_json <- Sys.getenv("ROPENDAL_GDRIVE_SECRET_JSON")
tokens_json <- Sys.getenv("ROPENDAL_GDRIVE_TOKENS_JSON", unset = file.path(dirname(secret_json), "tokens.json"))
gdrive_root <- Sys.getenv("ROPENDAL_GDRIVE_ROOT", unset = "Ropendal")
gdrive_file <- Sys.getenv("ROPENDAL_GDRIVE_FILE", unset = "map_catalog.txt")

drive_fs <- opendal(
  "gdrive",
  root = gdrive_root,
  auth = credentials_gdrive3(
    secret_json = secret_json,
    tokens_json = tokens_json,
    scope = "https://www.googleapis.com/auth/drive"
  )
)
drive_head <- rawToChar(fs_read(drive_fs, gdrive_file, size = 64))
drive_head
```

## Native C API roundtrip

The native API is for other R packages that want OpenDAL-backed byte I/O without
calling R while async work is running. A downstream package can declare
`LinkingTo: Ropendal`, include `<ropendal.h>`, submit async work, wait on Aio
handles, and read borrowed results or fill caller-owned buffers.

We exercise that installed C API in-process with
[`Rtinycc`](https://github.com/sounkou-bioinfo/Rtinycc). We assemble the C source
in pieces, use Rtinycc's bundled
[Protothreads](http://dunkels.com/adam/pt/) header, and let R resume a native
state machine while doing ordinary R-side work between resumptions. The `Rprintf()`
calls below are demo instrumentation executed only when R resumes the C task on
the main thread; OpenDAL background work does not call R.

```{r native-c-api-setup, eval = requireNamespace("Rtinycc", quietly = TRUE)}
root <- tempfile("ropendal-c-api-readme-")
dir.create(root, recursive = TRUE)

ropendal_lib <- list.files(
  system.file("libs", package = "Ropendal"),
  pattern = paste0("Ropendal", .Platform$dynlib.ext, "$"),
  recursive = TRUE,
  full.names = TRUE
)
```

We start with includes, status codes, and a native task struct. State that must
survive a protothread yield lives in the struct, not in local C variables.

```{r native-c-api-c-state, eval = requireNamespace("Rtinycc", quietly = TRUE)}
c_api_state <- r"(
#include <rtinycc/pt.h>
#include <stdint.h>
#include <stdio.h>
#include <stdlib.h>
#include <string.h>
#include <R_ext/Print.h>
#include "ropendal.h"

enum { ROPENDAL_DEMO_DONE = 0, ROPENDAL_DEMO_RUNNING = 1, ROPENDAL_DEMO_ERROR = -1 };
static const uint8_t demo_payload[] = "hello native api\n";

typedef struct ropendal_demo_task {
  struct pt pt;
  ropendal_fs_t *fs;
  ropendal_aio_t *aio;
  ropendal_error_t *err;
  ropendal_write_options_t write_opts;
  ropendal_read_options_t read_opts;
  ropendal_ls_options_t ls_opts;
  const ropendal_entry_t *entry;
  const ropendal_entry_t *entries;
  size_t nentries;
  size_t nread;
  uint8_t dst[64];
  int status;
  int failed;
  int done;
  int tick;
  const char *step;
  char message[256];
} ropendal_demo_task_t;
)"
```

Next we define cleanup and error helpers plus a constructor. `ropendal_fs_open()`
copies the root configuration during the call, so the R string does not need to
outlive the constructor.

```{r native-c-api-c-lifecycle, eval = requireNamespace("Rtinycc", quietly = TRUE)}
c_api_lifecycle <- r"(
static void demo_set_error(ropendal_demo_task_t *task, const char *fallback) {
  const char *message = fallback ? fallback : "native task failed";
  if (task && task->err) {
    const char *native_message = ropendal_error_message(task->err);
    if (native_message && native_message[0]) message = native_message;
  }
  if (task) {
    snprintf(task->message, sizeof(task->message), "%s", message);
    task->step = message;
    task->failed = 1;
    if (task->err) {
      ropendal_error_release(task->err);
      task->err = 0;
    }
  }
}

#define DEMO_FAIL(task, message) do { demo_set_error((task), (message)); PT_EXIT(&(task)->pt); } while (0)

void *ropendal_demo_open(const char *root) {
  ropendal_demo_task_t *task = (ropendal_demo_task_t *)calloc(1, sizeof(ropendal_demo_task_t));
  if (!task) return NULL;
  PT_INIT(&task->pt);
  task->step = "open filesystem";

  ropendal_kv_t cfg = { sizeof(ropendal_kv_t), "root", root };
  task->status = ropendal_fs_open("fs", &cfg, 1, &task->fs, &task->err);
  if (task->status != ROPENDAL_OK) demo_set_error(task, "ropendal_fs_open failed");
  else task->step = "filesystem ready";
  return task;
}

void ropendal_demo_free(void *ptr) {
  ropendal_demo_task_t *task = (ropendal_demo_task_t *)ptr;
  if (!task) return;
  if (task->aio) ropendal_aio_release(task->aio);
  if (task->fs) ropendal_fs_release(task->fs);
  if (task->err) ropendal_error_release(task->err);
  free(task);
}
)"
```

The protothread body submits async work, yields to R, polls for completion, and
then extracts results. This performs write, stat, listing, and a read into a
caller-owned C buffer.

```{r native-c-api-c-protothread, eval = requireNamespace("Rtinycc", quietly = TRUE)}
c_api_protothread <- r"(
static int ropendal_demo_resume_internal(ropendal_demo_task_t *task) {
  PT_BEGIN(&task->pt);

  task->step = "submit write";
  memset(&task->write_opts, 0, sizeof(task->write_opts));
  task->write_opts.struct_size = sizeof(task->write_opts);
  task->write_opts.path = "native.txt";
  task->status = ropendal_write_aio(
    task->fs, &task->write_opts, demo_payload, sizeof(demo_payload) - 1, &task->aio, &task->err
  );
  if (task->status != ROPENDAL_OK) DEMO_FAIL(task, "write submit failed");
  task->step = "wait write";
  PT_YIELD(&task->pt);
  while (ropendal_aio_poll(task->aio) == ROPENDAL_AIO_PENDING) PT_YIELD(&task->pt);
  task->status = ropendal_aio_wait(task->aio, -1, &task->err);
  if (task->status != ROPENDAL_OK) DEMO_FAIL(task, "write wait failed");
  ropendal_aio_release(task->aio);
  task->aio = 0;

  task->step = "submit stat";
  task->status = ropendal_stat_aio(task->fs, "native.txt", 0, 0, &task->aio, &task->err);
  if (task->status != ROPENDAL_OK) DEMO_FAIL(task, "stat submit failed");
  task->step = "wait stat";
  PT_YIELD(&task->pt);
  while (ropendal_aio_poll(task->aio) == ROPENDAL_AIO_PENDING) PT_YIELD(&task->pt);
  task->status = ropendal_aio_wait(task->aio, -1, &task->err);
  if (task->status != ROPENDAL_OK) DEMO_FAIL(task, "stat wait failed");
  task->status = ropendal_aio_result_entry(task->aio, &task->entry, &task->err);
  if (task->status != ROPENDAL_OK || !task->entry || !task->entry->has_content_length) {
    DEMO_FAIL(task, "stat result failed");
  }
  if (task->entry->content_length != sizeof(demo_payload) - 1) DEMO_FAIL(task, "unexpected stat size");
  ropendal_aio_release(task->aio);
  task->aio = 0;

  task->step = "submit list";
  memset(&task->ls_opts, 0, sizeof(task->ls_opts));
  task->ls_opts.struct_size = sizeof(task->ls_opts);
  task->ls_opts.path = "";
  task->status = ropendal_ls_aio(task->fs, &task->ls_opts, &task->aio, &task->err);
  if (task->status != ROPENDAL_OK) DEMO_FAIL(task, "ls submit failed");
  task->step = "wait list";
  PT_YIELD(&task->pt);
  while (ropendal_aio_poll(task->aio) == ROPENDAL_AIO_PENDING) PT_YIELD(&task->pt);
  task->status = ropendal_aio_wait(task->aio, -1, &task->err);
  if (task->status != ROPENDAL_OK) DEMO_FAIL(task, "ls wait failed");
  task->status = ropendal_aio_result_entries(task->aio, &task->entries, &task->nentries, &task->err);
  if (task->status != ROPENDAL_OK || task->nentries == 0) DEMO_FAIL(task, "ls result failed");
  ropendal_aio_release(task->aio);
  task->aio = 0;

  task->step = "submit read_into";
  memset(&task->read_opts, 0, sizeof(task->read_opts));
  memset(task->dst, 0, sizeof(task->dst));
  task->read_opts.struct_size = sizeof(task->read_opts);
  task->read_opts.path = "native.txt";
  task->status = ropendal_read_into_aio(task->fs, &task->read_opts, task->dst, sizeof(task->dst), &task->aio, &task->err);
  if (task->status != ROPENDAL_OK) DEMO_FAIL(task, "read_into submit failed");
  task->step = "wait read_into";
  PT_YIELD(&task->pt);
  while (ropendal_aio_poll(task->aio) == ROPENDAL_AIO_PENDING) PT_YIELD(&task->pt);
  task->status = ropendal_aio_wait(task->aio, -1, &task->err);
  if (task->status != ROPENDAL_OK) DEMO_FAIL(task, "read_into wait failed");
  task->status = ropendal_aio_result_nread(task->aio, &task->nread, &task->err);
  if (task->status != ROPENDAL_OK) DEMO_FAIL(task, "read_into nread failed");
  if (task->nread != sizeof(demo_payload) - 1 || memcmp(task->dst, demo_payload, task->nread) != 0) {
    DEMO_FAIL(task, "read_into bytes mismatch");
  }

  ropendal_aio_release(task->aio);
  task->aio = 0;
  task->step = "complete";
  task->done = 1;
  PT_END(&task->pt);
}
)"
```

Finally we expose small C entry points for Rtinycc to call.

```{r native-c-api-c-exports, eval = requireNamespace("Rtinycc", quietly = TRUE)}
c_api_exports <- r"(
static const char *ropendal_demo_status_name(int status) {
  if (status == ROPENDAL_DEMO_RUNNING) return "running";
  if (status == ROPENDAL_DEMO_DONE) return "done";
  return "error";
}

static int ropendal_demo_report(ropendal_demo_task_t *task, int status) {
  task->tick += 1;
  Rprintf(
    "native request %d: %s; %s\n",
    task->tick,
    ropendal_demo_status_name(status),
    task->step ? task->step : "unknown"
  );
  return status;
}

int ropendal_demo_resume(void *ptr) {
  ropendal_demo_task_t *task = (ropendal_demo_task_t *)ptr;
  if (!task) return ROPENDAL_DEMO_ERROR;
  if (task->failed) return ropendal_demo_report(task, ROPENDAL_DEMO_ERROR);
  if (task->done) return ropendal_demo_report(task, ROPENDAL_DEMO_DONE);

  task->status = ropendal_demo_resume_internal(task);
  if (task->failed) return ropendal_demo_report(task, ROPENDAL_DEMO_ERROR);
  if (task->status == PT_YIELDED || task->status == PT_WAITING) {
    return ropendal_demo_report(task, ROPENDAL_DEMO_RUNNING);
  }
  task->done = 1;
  return ropendal_demo_report(task, ROPENDAL_DEMO_DONE);
}

int ropendal_demo_nread(void *ptr) {
  ropendal_demo_task_t *task = (ropendal_demo_task_t *)ptr;
  return task ? (int)task->nread : -1;
}

const char *ropendal_demo_error(void *ptr) {
  ropendal_demo_task_t *task = (ropendal_demo_task_t *)ptr;
  return task ? task->message : "null task";
}

)"
```

We compile and bind that native code in memory.

```{r native-c-api-compile, eval = requireNamespace("Rtinycc", quietly = TRUE)}
c_api_code <- paste(c_api_state, c_api_lifecycle, c_api_protothread, c_api_exports, sep = "\n")

ffi <- Rtinycc::tcc_ffi() |>
  Rtinycc::tcc_include(system.file("include", package = "Rtinycc")) |>
  Rtinycc::tcc_include(system.file("include", package = "Ropendal")) |>
  Rtinycc::tcc_library(ropendal_lib[[1]]) |>
  Rtinycc::tcc_source(c_api_code) |>
  Rtinycc::tcc_bind(
    ropendal_demo_open = list(args = list("cstring"), returns = "ptr"),
    ropendal_demo_resume = list(args = list("ptr"), returns = "i32"),
    ropendal_demo_nread = list(args = list("ptr"), returns = "i32"),
    ropendal_demo_error = list(args = list("ptr"), returns = "cstring"),
    ropendal_demo_free = list(args = list("ptr"), returns = "void")
  ) |>
  Rtinycc::tcc_compile()
```

R controls the scheduling loop. Between native resumptions we do ordinary R work;
the native code never creates R objects while OpenDAL I/O is running.

```{r native-c-api-run, eval = requireNamespace("Rtinycc", quietly = TRUE)}
task <- ffi$ropendal_demo_open(root)
status <- 1L
r_ticks <- 0L
r_work <- 0L
while (status > 0L) {
  status <- ffi$ropendal_demo_resume(task)
  r_ticks <- r_ticks + 1L
  r_work <- r_work + sum(seq_len(1000L))
  Sys.sleep(0.001)
}
if (status < 0L) {
  message <- ffi$ropendal_demo_error(task)
  ffi$ropendal_demo_free(task)
  stop(message, call. = FALSE)
}
nread <- ffi$ropendal_demo_nread(task)
ffi$ropendal_demo_free(task)
c(
  r_ticks = r_ticks,
  r_work = r_work,
  bytes_read_into_c_buffer = nread
)
```

## Development

Common targets:

```{bash make-help, comment = ""}
make --no-print-directory help
```
