---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit this file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Ropendal

Ropendal is an abstract filesystem interface for R backed by the Rust crate of
[Apache OpenDAL](https://opendal.apache.org/). The package is byte-first:
operations move raw bytes, then explicitly materialize into R objects through modes
and serializers.

```{r badges, echo = FALSE}
# Keep this block intentionally minimal in README to avoid rendering noise.
```

## Quick start

```{r quick-start}
library(Ropendal)

root <- file.path(tempdir(), "ropendal-readme")
unlink(root, recursive = TRUE)
dir.create(root, recursive = TRUE)

fs <- opendal("fs", root = root)

fs_write(fs, "note.txt", charToRaw("hello ropendal\n"))
raw <- fs_read(fs, "note.txt")
raw

stat <- fs_stat(fs, "note.txt")
stat[c("path", "type", "size")]

entries <- fs_ls(fs)
vapply(entries, `[[`, character(1), "path")
```

## Async reads and explicit wait

```{r quick-async}
aio <- fs_read_aio(fs, "note.txt")
call_aio(aio)
collect_aio(aio)
```

## Documentation and concepts

The README now points to full vignette guidance.

- <https://sounkou-bioinfo.github.io/Ropendal/articles/getting-started.html>
- <https://sounkou-bioinfo.github.io/Ropendal/articles/abstract-filesystem.html>
- <https://sounkou-bioinfo.github.io/Ropendal/articles/async-aio.html>
- <https://sounkou-bioinfo.github.io/Ropendal/articles/serializers.html>
- <https://sounkou-bioinfo.github.io/Ropendal/articles/credentials.html>
- <https://sounkou-bioinfo.github.io/Ropendal/articles/native-c-api.html>
- <https://sounkou-bioinfo.github.io/Ropendal/articles/internals.html>

More implementation notes and completion status are tracked in
`design/STATUS.md`.

## Installation

Ropendal is available from R-universe:

```{r install-default, eval = FALSE}
install.packages(
  "Ropendal",
  repos = c("https://sounkou-bioinfo.r-universe.dev", "https://cloud.r-project.org")
)
```

Source builds can still use explicit OpenDAL feature flags when you need custom
provider wiring.

```{r install-source-features, eval = FALSE}
# Keep only local filesystem, HTTP, S3-compatible, and Google Drive support.
install.packages(
  "Ropendal",
  repos = c("https://sounkou-bioinfo.r-universe.dev", "https://cloud.r-project.org"),
  type = "source",
  configure.args = "--without-default-rust-features --with-rust-features=fs,http,s3,gdrive"
)

# Add the current cloud-service feature group explicitly.
install.packages(
  "Ropendal",
  repos = c("https://sounkou-bioinfo.r-universe.dev", "https://cloud.r-project.org"),
  type = "source",
  configure.args = "--enable-cloud"
)
```

## Development

Common targets:

```{bash make-help, comment = ""}
make --no-print-directory help
```
