---
title: "Native C API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Native C API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Ropendal exposes a pure C API in `inst/include/ropendal.h` for downstream native
packages. The header is generated from the same Rust surface used by the R API.

## Async-first surface

C calls mirror the package’s async behavior:

- `*_aio()` for submission,
- `ropendal_aio_wait()` / monitor helpers for completion,
- explicit error values for operation errors.

## Minimal usage pattern

```{r, eval = FALSE}
// allocate a filesystem handle
ropendal_fs_t *fs = NULL;
ropendal_error_t *err = NULL;
ropendal_fs_open("fs", NULL, 0, &fs, &err);

// submit async read and wait
ropendal_read_options_t opts = {
  .struct_size = sizeof(opts),
  .path = "object.bin"
};

ropendal_aio_t *aio = NULL;
ropendal_status_t status = ropendal_read_aio(fs, &opts, &aio, &err);
if (status == ROPENDAL_OK) {
  ropendal_aio_wait(aio, -1, &err);
  const uint8_t *data = NULL;
  size_t len = 0;
  ropendal_aio_result_bytes(aio, &data, &len, &err);
}

// always release native resources
ropendal_aio_release(aio);
ropendal_fs_release(fs);
ropendal_error_release(err);
```

## Why this exists

Downstream packages can perform byte-oriented async operations without going through
R-level raw allocation. This keeps the native path fast and explicit while using
the same transport semantics as the user-facing R API.
